---
title: "Model_Details"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model_Details}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette details the specific models used in this package, along with their parameters of interest, along with some of the relevant theory behind them.

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(dadnowcast)
```

In this package the assumed structure of the data is that there will be a response variable DAD ($y_t$) occurring over time $t$, and several explanatory variables RVDSS ($x_t$), PTSOS ($z_{1,t}$), and CNISP ($z_{2,t}$). $z_1$, and $z_2$ are assumed to be from the same data generating process and thus at least somewhat correlated, while $x$ is assumed to come from a different data generating process.

All of the models in this package take four arguments: Three are for data; `Y_train`, `X_train`, and `X_nowcast`, which represent the observed response variable, the correlating observed explanatory variables, and the observations of the explanatory variables that have no corresponding response variable observations which are used as a basis for the nowcasts created. The last is `params` which is for additional parameters for the the specific model.

# Linear Model

The first and simplest models that we will be looking at in this project are linear models where the value of DAD we are nowcasting is dependent on the current values of the explanatory variables:
$$
y_t = \alpha_0 x_t + \beta_1 z_{t,1} + \beta_2 z_{t,2} + \epsilon_t
$$

We can build on this basic model by adding the complexity of lagging the values of the explanatory variables:
$$
y_t = \alpha_0 x_t + \alpha_1 x_{t-1} + \beta_1 z_{t,1} + \beta_2 z_{t,2} + \epsilon_t
$$  

The function `fit_LM` fits a linear model with given parameters, there are no options for any other parameters in this model. This model is the simplest and easiest to run and interpret, however it has the drawbacks of not directly accounting for time or previous values of the response variable, and not accounting for non-linear patterns.

# AR

The function `fit_AR` fits a basic Auto-Regressive (AR) model on the response variable `Y_train`, and takes a parameter `n.ahead` to indicate how many predictions the function should make, as well as a parameter `p` which indicates the degree of the AR model. A higher value of `p` means that the model will directly consider more previous terms in the calculation, a high `p` can easily lead to overfitting depending on the data, since previous terms may not be strongly correlated with the current term.

This function also includes many parameter options to fit many more advanced variations of AR models, but these are for the convenience of future functions. An AR(p) model follows the form:

$$
y_t = \sum_{i=0}^p \beta_{t-i} y_{t-i}
$$

This is a simple AR model, which is only based on time and the previous trend of the response variable, it does not take any exogenous variables into account when fitting the model and making predictions.

## ARX

The first variation is the function `fit_ARX` which fits an AR model with exogenous factors. An ARX model follows the form:

$$
y_t = \sum_{i=0}^n \alpha_{t-i} x_{t-i} + \sum_{i=0}^p \gamma_{t-i} z_{t-i,1} + \sum_{i=0}^p \gamma_{t-i} z_{t-i,2} + \sum_{j=1}^m \beta_{t-i} y_{t-i} + \epsilon_t
$$

The function `fit_ARX` takes the additional `params` argument `p` which indicates the degree of the AR model. While it is still possible to use `n.ahead` in `fit_ARX`, it is unnecessary due to the function automatically using the length of `X_nowcast` as the number of predictions to make.

This model builds on the simple AR model by now taking into account the exogenous variables.

## ARIMAX

A natural extension of the AR model is to add the moving average (MA) and integration parts (I) to make a function that fits ARIMAX models. The function then follows the form:

$$
(1 - \sum^p_{i=1}\alpha_iL^{i}) X_t = (1 + \sum^q_{i=1}\theta_iL^{i})
$$

The function `fit_ARIMA` takes additional `params` arguments: 

- `p` which indicates the degree of the AR portion. A higher value of `p` means that the model will directly consider more previous terms in the calculation which can lead to overfitting depending on the data.
- `d` which indicates the degree of differencing. A high value for `d` means that the model requires more differencing to make it stationary, in general lower values of `d` are preferred since it is easy to ovefit a model with a large value.
- `q` which indicates the MA degree. A large value of `q` means that the moving average takes more previous terms into account and makes the moving average part of the model smoother while loosing some of the subtly in the movement of the data.

This model builds on an ARX model by incorporating a moving average and differencing in it, these allow for the model to capture more complex trends in the data, but increase the risk of overfitting or the model identifying trends that are not actually present in the data.

# GAM

The function `fit_GAM` fits generalized additive models, it takes the general form:

$$
y_t = \sum_{i=0}^n \alpha_i b_i(x_{t}) + \sum_{j=0}^m \beta_j b_j(z_{t,1}) + \sum_{k=0}^l \gamma_k b_k(z_{t,2}) + \epsilon_t
$$  

Where $b_i(), \ b_j(), \ b_k()$ are all basis splines.

The function `fit_GAM` fits a GAM function on the given data. There are multiple `params` options to specify parameters for the type of GAM function that is desired; 

* `family` specifies the distribution and link function for the model, it defaults to a gaussian distribution but there are many possible values to use, they are specified in `family`, and `family.mgcv`.
* `smooths` is a named list indicating which smoothing parameters to use on each variable.
* `bs` giving the basis smoothing function. Each of these parameters are optional and have appropriate built in defaults, but the parameters allow for a highly customizable function.

A GAM function is much more flexible than a linear model or an ARIMA model, and can capture complex patterns in the data, however it does not necessarily take the time bound nature of this data into account without modifications (ie using the lag terms). A GAM can also be prone to overfitting to the training data.

# Random Forest

The function `fit_RF` fits a random forest model on the given data using the package `randomForest`. This function uses the same data variables, `Y_train`, `X_train`, and `X_nowcast`. There are parameters:

- `ntree` which specifies the number of trees to grow the default is 500, a higher value leads to a lower bias as each individual tree has a lower impact on the final outcome, but at the expense of being more computationally intense.
- `mtry` is the number of variables used as candidates at each split the default value is the square root of the number of explanatory variables used in the model, a larger value of mtry makes the trees within the forest more correlated with each and when the value of mtry is equal to the number of parameters the model is just bagging.
- `weights` for weighting the sample data, this is used if the user wishes to emphasize parts of the data, but in general is not necessary.
- `replace` specifies if sampling should be done with or without replacement the default is TRUE. 
- The final two options are both for limiting the size of the trees grown which is not necessary to do manually but is a very nice option to have, especially if computing time is a concern. 
   - `maxnodes` which limits the number of terminal nodes that a tree can have which defaults to letting the tree grow without restriction. 
   
   - `nodesize` which sets the minimum size of the terminal nodes and defaults to 5.
   
A random forest is a very flexible model which is very good at finding complex relationships between many variables.

# XGBoost

The function `fit_XGBoost` useds the package `xgboost` to create an extreme gradient boosted model. This model builds a series of trees sequentially and each subsequent tree learns from the trees before it. This model has several parameter options:

 - `nrounds` this specifies the maximum number of boosting iterations. The function `fit_xgboost` will automatically tune this value to the data that it is given if there is no default value given.
 - `verbose` specifies if information about the model will be printed out when creating it, 0 makes it silent, the default is 1.
 - `XGBparams` is a list of more parameters used in the model, a full list of the options is available in `xgboost::xgb.params`, but the most important of them is `max_depth` which sets the depth of the trees used in the model. A higher `max_depth` means that the tree will be deeper allowing for more end nodes, and a higher potential for overfitting. This function will automatically tune the value of `max_depth` if one is not specified.

# Kalman Filter

The function `fit_KalmanFilter` uses the package `KFAS` to create a Kalman-Filter on the given data. A Kalman Filter works by creating a prediction and then updating it with real data when the real data is received. The combination of the prediction and real data is then used to create another prediction step. `fit_KalmanFilter` only has a parameter option for `degree` which specifies the trend component of the model, the default is 1 which is for local level, 2 makes it a local linear trend.
