---
title: "An R Package for Nowcasting DAD Data"
output: rmarkdown::html_vignette
author: "Devan Becker and Timothy Marsh"
vignette: >
  %\VignetteIndexEntry{Package_Details}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: Package_Details.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo=FALSE}
library(fatherStay)
```

## Objective

This package is designed to perform nowcasting on a particular set of data. The inputs are defined by the intended user of this package, the interface will be user-friendly for intermediate R programmers, and the outputs will provide useful insights into the performance of the nowcasting model.

This package expects the data to look something like the following:

| date       | $y$   | $x_1$  | $z_1$  | $z_2$  |
|------------|-----|-----|-----|-----|
| 2020-01-01 | 1.1 | 0.5 | 2.3 | 1.2 |
| 2020-01-02 | 1.5 | 0.2 | 2.7 | 1.2 |
| 2020-01-03 | NA  | 0.1 | 2.8 | 1.5 |
| 2020-01-04 | 1.5 | NA  | 2.7 | 1.3 |
| 2020-01-05 | 1.8 | 0.2 | 2.9 | 1.2 |
| ...        | ... | ... | ... | ... |
| 2020-12-26 | 2.5 | 1.2 | 1.7 | 2.2 |
| 2020-12-27 | 2.9 | 0.8 | 0.7 | 2.5 |
| 2020-12-28 | NA  | 0.2 | 2.7 | 2.9 |
| 2020-12-29 | NA  | 0.6 | 2.2 | 1.9 |
| 2020-12-30 | NA  | 0.4 | 2.0 | 3.9 |
| 2020-12-31 | NA  | 0.2 | 1.0 | 3.8 |

The string of trailing NA values in $y$ are interpreted by the package as the values to be nowcasted. Note that the values of $x_1$, $z_1$, and $z_2$ are available at these times, even though the values of $y$ are not.

The `date` should be converted into a datetime object in R prior to use in this package, e.g. with `lubridate::ymd()`. Note that the column names can be arbitrary.

Note the presence of NA values in the response and the predictors prior to nowcasting, which must be imputed for many of the methods we propose. Helper functions can be created to do some of the data processing steps as needed.

## Nowcasting

Starting with defining the specific problem that we are wanting to solve. When referring to nowcasting we denote the measurement of DAD at time $t$ by the value $Y_t$ and the measurements of the explanatory variables at time $t$ to be the values $x_{1,t}$ for RVDSS and $z_{j,t}$ where $j \in \{1,2\}$ representing CNISP, and PTSOS. Then the problem that we are working is how to predict the value of $Y_t$ given the values $x_{1,t}$, $z_{j,t}$, and previous (lag) values for each variable (including $y_{t-l}$).

We will be making several key assumptions in this project. The first is that any missing data for the explanatory variables can be at least somewhat successfully imputed by reasonable estimates. We will also assume that we have access to all the updated values of the explanatory variables at the time $t$ when we are making a prediction for $y_t$ (or they can be reasonably well imputed).

## Uncertainty Quantification

Regardless of the model used, the goal of this package is to predict the trailing NA values in the response variable and to provide estimates of the uncertainty of these predictions. Some of the proposed models include uncertainty estimates as part of their machinery (for example, linear models have closed-form prediction intervals), but others require numerical techniques.

A model-agnostic estimate of prediction uncertainty is Ensemble-based Batch Prediction Intervals (EnBPI) [@xuConformalPredictionInterval2021]. Suppose we're doing 5-step ahead prediction with a batch size of 50. We start by randomly sampling a "batch" from our complete data (not including the data to be nowcasted), that is, we take a string of 50 observations as well as the 5 next data points. We fit the model to the 50 observations and we make predictions for the next 5 data points. We record how far away our prediction is from the observed data. We do this many times to build up a data set of 5-step ahead prediction errors. We can look at, say, the 2.5\% and 97.5\% quantiles of these errors to get the width of our prediction interval. When we actually perform the nowcasts on the missing data, we center these intervals on our predictions.

Since this approach works for any model (regardless of whether there's a theoretical or closed-form solution for prediction intervals), we will apply this to all models for consistency. The only exception will be for any Bayesian models we might implement, for which we'll use the posterior of the predicted values for all numerical summaries.

## A Note on Predictions

For models that use lagged values of DAD, special considerations must be made. In order to predict two steps ahead, the one-step ahead prediction is required. If the one-step-ahead prediction is poor, then the two-step-ahead prediction will also be poor. In this way, the k-step-ahead prediction gets more variable as k increases.

This is not the case for models that do not rely on lagged values of DAD; predictions for k-steps-ahead can simply be made based on the covariates (there is no distinction for lagged versus current values of the covariates so long as the information is available). However, since models with lagged values of the response incorporate more information, they can generally be expected to make better predictions.

In this package, any method that is implemented must have a way to handle $k$-step ahead predictions.

### Prediction of Unobserved Locations

The models will only be fit to data from a given location, but it may be reasonable to use the estimated model from one location to predict another location, for example in a location for which DAD is never observed.

For non-lag-based models, this is straightforward. However, for models that use the lags of the response, any errors in the predictions for the first few values of DAD will propagate throughout all later predictions.

The model will be designed to allow the user to input the formula, then the resultant object can be used for predictions on arbitrary data. This allows fitted models to be used to predict any other location. If the user wants to make predictions for locations with no DAD recorded, we make the following suggestions:

1. Make predictions for locations where DAD *is* known, and observe the prediction errors. 
2. There is no restriction in the package that DAD must be the response. Perhaps try fitting a model with, say, PTSOS as the response and predict the PTSOS at the other locations, observing the prediction errors. This gives a glimpse into how well one location's data can predict another, especially given that PTSOS and CNISP are measuring something very similar to the response.

The prediction errors from other locations should be interpreted carefully since there is absolutely no guarantee that the relationships in one location transfer to another.

## Imputation

We expect there to be some missing data in the predictor variables. We plan on using imputation to deal with them since many time series methods require there to be no gaps in the data. The goal is that we will have relatively simple methods to impute the values of the explanatory variables $x_1$, $z_1$, $z_2$, and if necessary $y_t$, and the package will allow the user to specify the type of imputation to be used. The data has multiple potentially correlated variables, this could be used to improve the imputations, this will come in useful when we need to impute large sections of the explanatory variables such as when a region is missing a certain measurement, however for imputing $y_t$ we will use simpler methods since imputation is not the main focus of this project.

There are several existing R packages that allow for imputation and work in a time series context, and while we do not want the program we are developing to be dependent on many packages it makes sense to use an existing package for the imputation since it is not a focus of the project. 

The package that we will be using in this project is imputeTS, a package in R for imputing univariate time series. It has the advantage of being specifically for time series data. It features a variety of complex methods for imputation such as; Kalman smoothing, and interpolation, as well as simple methods like using the last observation or the mean [@moritzImputeTSTimeSeries2017]. While this package does not take advantage of the multivariate nature of our data it is still a good option to impute the missing values in our data.

## Potential Models

There are several different techniques for nowcasting in a time series context, in general they are organized from simplest to more complex:

### Simple Linear Models

The first and simplest models that we will be looking at in this project are linear models where the value of DAD we are nowcasting is dependent on the current values of the explanatory variables:
$$
y_t = \alpha_0 x_t + \beta_1 z_{t,1} + \beta_2 z_{t,2} + \epsilon_t
$$

We can build on this basic model by adding the complexity of lagging the values of the explanatory variables:
$$
y_t = \alpha_0 x_t + \alpha_1 x_{t-1} + \beta_1 z_{t,1} + \beta_2 z_{t,2} + \epsilon_t
$$

### Mechanistic Models

In the simple linear models, the user can input whatever predictors they want (including lag terms and non-linear transformations). However, the nature of the data suggests a possible mechanistic model. The following model structure was suggested by David Champredon.

Let $D_t$ be the DAD admission data ($y_t$ in the previous notation), $R_t$ be the positive tests from the RVDSS data ($x_t$), and $C_t$ and $P_t$ be the total admissions and positive tests from the CNISP and PTSOS data, respectively ($z_1$ and $z_2$). The following assumptions lay the foundation of the mechanistic model:

1. The total admissions are a constant fraction of the total positive tests.
    - $D_t \approx \eta R_t$, e.g. $\eta$ = 20% of cases are hospitalized.
    - $\eta$ is the "virulence"
2. The total admissions are a constant fraction of *each of* the provincial and national hospital data.
    - $\hat s_c D_t \approx C_t$ and $\hat s_p D_t \approx P_t$
    - $\hat s_c$ and $\hat s_p$ are the sampling rates, which are assumed to be known from historical data.

This gives us three equations with parameters/error terms to be estimated simultaneously:
\begin{align*}
E(D_t) &= \eta R_t\\
E(C_t) &= \eta \hat s_c R_t\\
E(P_t) &= \eta \hat s_p R_t
\end{align*}

The interpretation of this model is that the test positivity is linearly related to hopsitalizations, with a coefficient that is consistent across DAD, PTSOS, and CNISP. There is an adjustment to the coefficients for CNISP and PTSOS as they are assumed to be samples from DAD. The benefit of this framework is a more robust estimate of $\eta$, but this requires fairly strict assumptions on the data generating process.

Assuming independence between $D_t$, $C_t$, and $P_t$ (conditional on $R_t$ and the parameters), this model can be estimated via Bayesian methods or with MLE. The likelhood factors into three parts:
\begin{align*}
f(D_t, C_t, P_t | \eta, R_t, ...) &= f_d(D_t|\eta, R_t, ...) f_c(C_t|\eta, R_t, ...) f_p(P_t|\eta, R_t, ...)
\end{align*}
and we can assume an appropriate likelihood for each response. The "$...$" in each factor allows for likelihoods with further parameters. If we assume a normal distribution (i.e., simple linear regression for each), then the "$...$" takes the place of the variance terms (which could also allow for correlation). A Poisson GLM would not require further parameters, but a Negative Binomial GLM would extend the Poisson GLM to include a dispersion parameter. There is no restriction that each model have the same distribution, but doing so may allow for a closed-form solution for $\eta$ by minimizing the log likelihood.

Some other notes:

- It may be beneficial to allow for correlated errors across the three models. It seems likely that if, say, CNISP is being underpredicted by $R_t$ then so is PTSOS. This is possible in a basic linear regression (Gaussian) framework, but difficult in GLMs in general. For Poisson and Negative Binomial, further research (as part of a separate project) into copula modelling or shared random effects may be appropriate.
- This model assumes that $\hat s_c$ and $\hat s_p$ are known from historical data, but since we have multiple response variables it would be feasible to estimate these from the data. The package can be modified to allow for this in the future.


### Non-linear Models

In additional all the previous linear models we will also be considering non-linear models/processes. In this project we are seeking to predict the current values of DAD, not create an explanatory model. This allows us to use many ML models since they are focused on creating predictions not on the explanatory power of the resulting model. 

The simplest option is to use splines to create a basic spline model of the response (DAD) based on the current values of our explanatory variables:
$$
y_t = \sum_{i=0}^n \alpha_i b_i(x_{t}) + \sum_{j=0}^m \beta_j b_j(z_{t,1}) + \sum_{k=0}^l \gamma_k b_k(z_{t,2}) + \epsilon_t
$$
Where $b_i(), \ b_j(), \ b_k()$ are all basis splines. There is room to expand this to include the past values of the explanatory variables.

Random forests are also a good technique for us to use on this project, since they are relatively easy to implement and are a powerful tool, the predictive capabilities can be further enhanced by using techniques like XGboost. Random Forests work best when there are many variables for them to select, and find non-linear patterns between. In this project, we only have our three main covariates to work with, however the time bound nature of our data allows us to use the lagged values of the covariates, to enhance the number of variables we are working with. The Random Forest model would then be able to predict future values of DAD based on some combination of the current and lagged values of the explanatory variables. There are also many available R packages for Gaussian process regression which may be applicable to these data [@rasmussenGaussianProcessesMachine2008]. 

There are other potential models that we may want to use in this project such as simple neural nets [@richardsonNowcastingGDPUsing2019]. The models discussed here are the ones we will definitely be using and a good starting point, but throughout the project other useful models will likely come up and we will be using those as well. We will be careful to ensure that any models we apply are well-established models with known properties and robust R packages; we won't reinvent the wheel but we also won't use models with unclear properties or error-ridden implementations.

The package will also provide an API to allow additional models. The user can specify a function (or functions) with well-defined (and well-documented) inputs and outputs. 

### ARX

A natural next step is to model the process as an ARX process, that is as an AR system that is based on previous values of the variables itself but also dependent on current and past values of some external explanatory variable. We can represent it by the following equations [@picciARXModelingTime2024]:

$$
y_t = \sum_{i=0}^n \alpha_{t-i} x_{t-i} + \sum_{i=0}^p \gamma_{t-i} z_{t-i,1} + \sum_{i=0}^p \gamma_{t-i} z_{t-i,2} + \sum_{j=1}^m \beta_{t-i} y_{t-i} + \epsilon_t
$$

Where $\alpha_i$ represents the impacts of the external variables and $\beta_i$'s are the parameters for the AR model and have the same interpretation of being measures of the impact of previous terms in the sequence. Note that the summation index for $y_{t-i}$ starts at 1 since the previous value will be used, but the sum starts at 0 for the covariates since we are using the current value of the covariates.

The ARX model is a special case of the ARIMAX model where the integration and moving average parts are both 0. The `arima()` function in R will be used for estimation, so expanding the ARX model to an ARIMAX one is straightforward.

### Kalman filter/Smoothing

A common method in nowcasting time series data is using the Kalman Filter [@rossiKalmanFilteringBased2015]. The Kalman filter predicts one step ahead and its algorithm works in two steps, first creating a prediction, and then after the actual data is observed, updating that prediction using the Kalman gain. This acts as a smoother, where the predictions can be treated as if they are equally as "true" as the observed values. In other words, this is treating the response as if it is measured with error and the true value is somewhere between the observed value and the prediction. Of course, this assumes that we are able to create good predictions, which is why Kalman filter models are often complex, including careful estimation of the covariance between the predictions and the observed values and often employing multivariate Gaussian Process modelling.

There are several R packages that can be used to run a Kalman Filter with varying levels of complexity [@tusellKalmanFiltering2011].

# References
