---
title: "Package_Details"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Package_Details}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(fatherStay)
```

## Objective

Our data are expected to look something like the following:

| date | y   | x1  | z1  | z2  |
|------|-----|-----|-----|-----|
| 1    | 1.1 | 0.5 | 2.3 | 1.2 |
| 2    | 1.5 | 0.2 | 2.7 | 1.2 |
| 3    | NA  | 0.1 | 2.8 | 1.5 |
| 4    | 1.5 | NA  | 2.7 | 1.3 |
| 5    | 1.8 | 0.2 | 2.9 | 1.2 |
| ...  | ... | ... | ... | ... |
| 204  | 2.5 | 1.2 | 1.7 | 2.2 |
| 205  | 2.9 | 0.8 | 0.7 | 2.5 |
| 206  | NA  | 0.2 | 2.7 | 2.9 |
| 207  | NA  | 0.6 | 2.2 | 1.9 |
| 209  | NA  | 0.4 | 2.0 | 3.9 |
| 210  | NA  | 0.2 | 1.0 | 3.8 |

Note the string of NA values in y in the last days of the study, which are the values to be nowcasted. Note also the presence of NA values in the response and the predictors.

## Imputation

We expect there to be some missing data in the predictor variables. Imputation is the best way to deal with them.



